On the spatial aggregation of ecosystem condition indicators

### 2025-01-22 manuscript/manuscript.qmd
more writing

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

@@

Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable.
However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
Without a good enough understand of the pitfalls in spatial aggregation of information, is can also lead to unintended and undetected systematic bias, as well as downright logical errors.
Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (ref), as well as to the spatial aggregation of relatively unmodified or raw variables, there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
We pay specific attention to the phenomenon of aggregation bias introduce by @rastetter1992.

@@

@@

@@

| Term             | Definition |
|------------------|------------|
| Metric           |            |
| Variable         |            |
| Indicator        |            |
| Reference levels |            |
| Scaling          |            |
| Transformation   |            |
| Truncation       |            |
| Normalisation    |            |
| Compensation     |            |
|                  |            |

@@

@@

# Why do we normalise? {#sec-norm}

@@

The spatial aggregation of data can lead to biases arising from a number of sources.
For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
In this paper, however, we are mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
They showed that when we take the average of two variables, each of which have associated attribute values (e.g. indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.

## Aggregation error and discrepancies between variables and indicators

@@

Another cause of potential errors or confusion arising during the spatial aggregation of indicators, that may not relate directly to the aggregation bias problem, has to do with deciding on the appropriate area weights to use.
The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is commonly agreed upon as desirable.
This has to do with the for appropriate data to inform the area weighting in the spatial aggregation.
Consider an indicator originating from point-based field surveys.
The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
(something here about indicator versions and good documentation)

## Aggregating to account for spatial autocorrelation

@@

Indicator developer subjectivity is a big factor as the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
This paper is aimed at providing some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
This will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.

@@

Something about sometimes spatial aggregation is an essential part of the indicator design ---

In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e. horizontal) aggregation only.

@@

Pathway 1 involves early normalisation suing the perhaps most common order of the three steps scaling, truncating and transforming.
The three steps may be done simultaneously as well.
Then the indicator is spatially aggregated.
Pathway 2 involves aggregating the variable before normalising.
Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised.
Pathway 4 illustrated the aggregation of a variable, with no normalisation.
This is the pathway commonly used for variable accounts in the SEE EA. Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
Instead one report the number of water bodies in each condition class (spatial statistics).
Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see @sec-pti).

@@

@@

This example is from Norway, where wolves are extinct in most of the country.
Therefore, most carnivore management areas will get an indicator value of 0.
Wolves exists i designated wolf zones in south-east Norway, although in small numbers.
In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.

\*Conclusion\*: Indicators should make intuitive sense to the readers.
However, there are sometimes good reasons for making things complicated.
In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explain just one time, and not for each indicator specifically.
It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
Wolf numbers could in this example be converted to wolf density.

@@

This example illustrates a variable normalised at plot level or aggregated to regions first.
At what scale is the ref value intended?

@@

The special thing about the billberry indicator is the different spatial resolution for the variable and for the reference level.
The variable is recorded at the scale of vegetation plots.
The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
Because the normalisation includes a truncation step, scaling the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).
Scaling at the plot scale would require require unique reference levels for each forest type.
However, both pathway 1 and 3 has been used for this variable in two different forest ecosystem condition assessments.
Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
This cause a general confusion about how how the same variable (and the same date) can produce different indicator values.

*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3, and spatially aggregated the variable, from plot scale to regional scale, before normalising the variable and potentially aggregating it further.
It would also be possible to use pathway 6.

The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
Often indicators and variables are both included side-by-side in ecosystem condition assessments/accounts.
This is also recommended practice following the SEEA EA. However, if the indicators that are presented are normalised before any spatial aggregation, then the interpretation of the indicators is not simply as a normalised version of the variable.
Indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, or some other aspect where values above and below the average value are able to compensate for each other.

@@

In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale.
When performing an ECA, one has the option to either aggregate variable value for each region (each ecosystem asset), and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons, and then aggregate the indicator values to the regional level (pathway 1).

@@

Separate assessments should give identical indicator values for the same areas, given the same underlying data.
When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments.
This was also the case in the billberry example, but here we used another example to illustrate the point further.

@@

# Recommendation

>>>>

keywords: 
  - SEEA EA
  - ecosystem condition
  - ecosystem accounting
  - indicators
  - aggregation bias
  - upscaling
date: last-modified
bibliography: bibliography.bib
format:
  elsevier-pdf:
    keep-tex: true
    journal:
      name: ---
      formatting: preprint
      model: 3p
      cite-style: authoryear
    include-in-header:
      text: '\usepackage{lineno}\linenumbers'
execute:
  cache: true
  warning: false
  message: false
editor: 
  markdown: 
    wrap: sentence
---

@@

Ecosystem accounting (EA), and other types of environmental assessments, are generally aimed at communicating major changes and trends in the extent and condition of ecosystems at a scale that is relevant for decision makers.
For ecosystem condition accounting (ECA) and assessments, this often requires that **metrics**[^1] used to describe the condition of ecosystems are aggregated spatially to produce a single value for a larger area.
This represents a type of knowledge synthesis, aimed at condensing the complexity of the real world into something actionable at a political level.
However, spatial aggregation can and will often change the interpretation of a metric and the descriptive properties and normative claims one can make based on it.
Without a good enough understand of the pitfalls in spatial aggregation of information, is can also lead to unintended and undetected systematic bias, as well as downright logical errors.
Although much attention has been devoted to the vertical aggregation of environmental indicators into indices (e.g. @langhans2014), as well as to the spatial aggregation of relatively unmodified or raw variables (e.g @allain2018) , there is to our knowledge no explicit deliberation on the special case of spatially aggregating highly modified indicators used for ecosystem condition accounting.
This paper therefore focus on explaining and providing guidelines on the spatial aggregation inherent to the System of Environmental Economic Accounting - Ecosystem Accounting (SEEA EA) - the most used ecosystem accounting (natural resource accounting) framework in the world (ref).
We pay specific attention to the phenomenon of aggregation bias [@rastetter1992] and on the different ways to interpret the information in ecosystem condition variables versus indicators.

@@

@stein2001

@finke2002

<https://www.sciencedirect.com/science/article/pii/S0016706198000779?casa_token=UTwC6C3DthYAAAAA:ZZ6eao2vlMClkMtTrdA043xgkyjfjwjpLsitDzfohdqiB3GvVPyQoDFzY1jgu4o176jpDjbWCZE>

read bib in knotters...

### Other types of aggregation not considered in this paper

Note that it is common in this field to use the term aggregation about the process of aggregating normalised indicators into indices (i.e. thematic or vertical aggregation; @langhans2014), but in this paper we are discussing spatial (i.e. horizontal) aggregation only.
The term spatial aggregation is also sometimes used to refer to **upscaling** in the sense of taking local information and making it applicable of translate it to larger, e.g. landscape or regional, scales [@knotters2010].
This process often involves predicting values across larger areas or surfaces, generally through models, including krieging-based models and distribution models.
When the data (e.g. indicator values) are made wall-to-wall, one can then more easily get point estimates for larger areas via simple aggregation functions such as summation or by taking the mean.
Using models to predict variable or indicator values is allowed under the SEEA EA. However, in this paper we are addressing the second part of this process only: of simplifying the spatial variation, from many to few data points.

@@

In ECAs, data are often not wall-to-wall, and when upscaling is not done, as defined above, the spatial aggregation needs to account for the spatial structure of the data in a different way, such as design-based approached (relying on random or balanced samples across the entire EEA), or a step-wise aggregation via ecosystem assets, homogenous ecosystem areas [@vallecillo2022] or homogenous impact areas [@kolstad_making_2024].
Many aggregation functions are available, such as sum, worst-rule, different percentiles, arithmetic, geometric or harmonic means.
When using means, area weighting is required when following SEEA EA.

@@

| Term                | Definition |
|---------------------|------------|
| Metric              |            |
| Variable            |            |
| Indicator           |            |
| Reference levels    |            |
| Scaling             |            |
| Transformation      |            |
| Truncation          |            |
| Normalisation       |            |
| Compensation        |            |
| Aggregation pathway |            |

@@

### Accounting for spatial autocorrelation {#-sec-auto}

One reason why we might want to spatially aggregate ecosystem condition metrics is to fix issues with spatial auto-correlation, that is to remove problematic spatial structure in the data.
For example if data is collected opportunistically with sampling size varying between areas, then performing partial spatial aggregation to some intermediate size areaa will give equal weight or importance to to all areas regardless of how many data points are included in each.

@@

# Why do we normalise variables? {#sec-norm}

@@

The spatial aggregation of data can lead to biases arising from a number of sources.
For example, direct extrapolation based on uneven or spatially biased sampling leads to biased estimates [@kolstad_making_2024].
In this paper, however, we are mostly interested in what we consider an extended case of aggregation error, sensu @rastetter1992.
They showed that when we take the average of two variables, each of which have associated attribute values (e.g. indicator values), then the attribute of the average variable value differs from the average of the attributes of the original variables.
This displacement is refereed to as aggregation error, and its effect become more pronounced when the non-linearity is strong between variables and indicators.
In the statistical standard for EAs and the recommended guidelines for ecosystem accounts [@unitednations2021; @unitednations2022], there is no mention of aggregation bias, and there seems to be little awareness in general of the ramification of choosing the wrong spatial aggregation method for variables.
We now describe two ways that aggregation bias affects ECAs: by causing an increasing divergence in the type of information held in ecosystem condition variables versus indicators, and secondly, how it can cause confusion about the effects of chosing different **aggregation pathways** during indicator development.

## Discrepancies between variables and indicators

@@

# Compliance across nested assessments

Another cause of potential errors or confusion arising during the spatial aggregation of indicators has to do with the number of aggregation levels.
The SEEA EA uses a principle of commensurability, meaning that the order of certain operations should not affect the resulting estimate or value.
Spesifically, it should not matter whether you aggregated vertically first (to create an index) and then horizontally (i.e. spatial), or the other way around.
Note that this prisiple may not always hold for other cases, such as aggregating across nested areas, e.g. from ecosystem assets to EAAs, although this is perhaps considered desirable.
This has to do with the use of area weighting in the spatial aggregation.
Consider an indicator originating from point-based field surveys.
The sampling location are selected using a stratified protocol that ensures a representative sample of similar size (same *n*) is made for each ecosystem asset.
If the variable is normalised early in the aggregation pathway to create indicator values for each locality, then the mean of indicator values for each region will be representative for those regions, and no area weighting is required.
For the next spatial level, when information from all ecosystem assets should be used to create an estimate for the EAA, ecosystem assets with large extents should weight more in the calculations (following the SEEA EA prinsiples of area weighting).
Note that one cannot skip this first level of spatial aggregation via ecosystem assets, since this is an integral part of the sampling design, leading to spatial autocorrelation (se @sec-auto).
However, if in a later ECA, ecosystem assets are defined as subsets of the initial assets (a finer scale account), then the same indicator, when aggregated to the EAA level, will have a different value.
A similar case would be is sampling was random and dispersed across the EAA, lets say a country, and indicator values were first calculated as some average across all sampling points, and then in a later account or assessment, the same indicator was estimated for separate regions.
When aggregating the indicator values from regions to the whole country, using area weighting, the national estimate will differ, albeit slightly, from the initial national estimate that did not include this step-wise aggregation.
Cases when two accounts or assessments use what is considered the same indicator but find different values for the same areas, is cause for both confusion, and potentially mistrust.
In our experience this can also lead to time consuming detective work, trying to find the causes for such discrepancies.
We suggest two immediate solution to this issue.
First, to find better ways to document the aggregation pathways that are being used for each indicator, as part of the indicator metadata.
Second, to find ways to distinguish between apparently identical indicators, for example using different version numbers or IDs.
We provide practical solution to both these suggestions in @sec-rec.

@@

Indicator developer subjectivity affects ECAs, in part because the process of designing indicators and estimating indicator values often left entirely to the them, with little real involvement from indicator users.
It is therefore important to be aware of the role that indicator developers have in shaping the type of information, and the narratives, that is presented to decision makers.
This paper is aimed at providing some recommendations for how to choose the right spatial aggregation method for variables and indicators, and how to report these choices.
This will make it easier for indicator developers to avoid some of the pitfalls from having a too casual approach to the choice of aggregation pathways.

@@

@@

Pathway 1 involves early normalisation using the perhaps most common order of the three steps scaling, truncating and transforming.
The three steps may be done simultaneously as well.
Then the indicator is spatially aggregated.
Pathway 2 involves aggregating the variable before normalising.
Pathway 3 is similar, but here there are two aggregation steps: one before and one after normalisation.
This can for example be the case when variables are aggregated to the scale corresponding to the reference levels before they can be normalised.
Pathway 4 illustrated the aggregation of a variable, with no normalisation.
This is the pathway commonly used for variable accounts in the SEE EA. Pathway 5 is a common pathway in the WFD, which does not include spatial aggregation of indicator values.
Instead one report the number of water bodies in each condition class (spatial statistics).
Pathway 6 is a suggested pathway for handling WFD indicator in ECAs (see @sec-pti).

@@

# Examples

@@

This example is from Norway, where wolves are extinct in most of the country.
Therefore, most carnivore management areas will get an indicator value of 0.
Wolves exist in designated wolf zones in south-east Norway, although in small numbers.
In the Norwegian Nature Index [@jakobsson2020], the variable follows pathway 1 and is normalised at the CMA level.
When aggregating the indicator to a national value, it is common for the general public to interpret this value as the status for wolves in Norway, and to connect this value to Norway's international obligations to maintain wolf numbers.
But note that even an unlimited number of wolves in the wolf zone would not to bring the national indicator value up to a level indicator good condition.
This is because the normalisation process truncates the variable, and does not allow overshooting values to compensate for lower values in the rest of the country.

\*Conclusion\*: Indicators should make intuitive sense to the readers.
However, there are sometimes good reasons for making things complicated.
In this example, The Norwegian Nature Index uses the same pathway for all indicators, so this common misunderstandings about what indicator values represents can be explaind just one time, and not for each indicator specifically.
It can also be a good idea to sometime go back to the variable, and see if it can be defined differently to avoid these kinds of confusions.
Wolf numbers could in this example be converted to wolf density.

@@

@@

The billberry indicator exemplifies different spatial resolutions for the variable and for the reference levels.
The variable is recorded at the scale of vegetation plots.
The reference level however, is designed with a regional spatial scale in mind, especially because of how it encompasses the known variation in forest types in a region to estimate the mean billberry coverage under the reference condition.
Because the normalisation includes a truncation step, scaling the variable at the plot scale (pathway 1; @fig-agg1) would not allow overshooting values to compensate for lower values when aggregating regional indicator values, which would then be negatively displaced ($\bar i$) relative to when aggregating variable values ($\hat i$; @fig-agg1).
Scaling at the plot scale would require require unique reference levels for each forest type.
However, both pathway 1 and 3 has been used for this variable in two different forest ecosystem condition assessments.
Due to non-negligeable truncation of the variable during the normalisation, the two indicators were subject to different levels of aggregation bias, and consequently the spatially aggregated indicator values for the two assessments became different.
This caused a general confusion about how how the same variable (and the same data) can produce different indicator values.

*Conclusion*: Given that the reference value is the way it is, this indicator should have followed pathway 3, and spatially aggregated the variable from plot scale to regional scale before normalising the variable and potentially aggregating it further.
It would also be possible to use pathway 6.

The wolves example also highlight the importance of making clear the differences in the interpretation of variables and indicators.
Indicators and variables are sometimes presented side-by-side in ecosystem condition assessments or accounts.
This is also recommended <!--# is this true? --> practice following the SEEA EA. However, due to aggregation bias, spatially aggregated indicators should not be interpreted as normalised versions of aggregated variables.
Indicators typically now reflect the average ecosystem condition, where areas in very good or very poor condition have limited ability to compensate for opposite extreme values elsewhere.
The spatially aggregated variable on the other hand, may reflect something like a sum of individuals, an average of a population, or some other aspect where values above and below the reference levels are able to compensate for each other.

@@

In this example, the reference levels are uniform throughout the EAA, and the normalisation includes a transformation in the form of a piecewise-linear transformation by anchoring X~60~ to 0.6 on the indicator scale.
When performing an ECA, one has the option to either aggregate variable values for each region (each ecosystem asset), and then normalise that value (pathway 2 and 3), or to normalise the variable at the scale of the original nature type polygons, and then aggregate the indicator values to the regional level (pathway 1).

@@

<!--# Separate assessments should give identical indicator values for the same areas, given the same underlying data. When this is not true it raises concerns about reproducibility and sacrifices the credibility of both assessments. This was also the case in the billberry example, but here we used another example to illustrate the point further. -->

@@

# Recommendation {#sec-rec}
